Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-06 kappa.tex
Add field area description first shot.

% ----------------------------------------------------------------------
\subsection{Geographic setting}
% ----------------------------------------------------------------------

% fig:locmap
\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{map-northamerica}}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas covered by the Cordilleran (CIS), Laurentide (LIS), Innuitian
           (IIS) and Greenland (GIS) ice sheets during the last glacial maximum
           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP][]{Dyke.2004}.
           The rectangular box delimits the model domain used throughout this
           thesis and on which most figures were drawn. The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets on the map.}}
  \label{fig:locmap}
\end{figure}

@@

@@

>>>>

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes. The Laurentide ice
sheet, largest of the three ice sheets, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
the Western Cordillera, our region of focus in the present thesis.

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{map-northamerica}}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas covered by the Cordilleran (CIS), Laurentide (LIS), Innuitian
           (IIS) and Greenland (GIS) ice sheets during the last glacial maximum
           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP][]{Dyke.2004}.
           The rectangular box delimits the model domain used throughout this
           thesis and on which most figures were drawn. The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets on the map.}}
  \label{fig:locmap}
\end{figure}

% ----------------------------------------------------------------------
\subsection{Topographic setting}
% ----------------------------------------------------------------------

In North American, the Western Cordillera forms a broad mountain system with an
average east-west width of a about 1000\,km. Its topography is a complex
arrangement of mountain ranges and intramontane basins. It was formed during
multiple accretion phases associated with the closure of several back-arc
oceanic trenches over a 200\,Ma-long geologic history.

Most of the major reliefs are found along two major bands that run parallel to
each other along the general orientation of the Cordillera. The western band
consists of the Pacific Coast orogenic belt. Major ice-covered reliefs include,
from North to South, the Alaska Range, the Wranglell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern band consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
rectilinear Rocky Mountain / Tintina Trench. It includes the Selwyn and Mac
Kenzie Mountains to the north, and the Rocky Mountains to the South. The brooks
range, in northern Alaska, has been glaciated but its glaciers were not
connected to the main body of the Cordilleran ice sheet.

The Cordillera comports several intermontane bansins, which, as we will see
further, host particular climate properties and significantly impacted patterns
of ice sheet flow. The two major depressions are the Interior Plateau of
central British Columbia, and the Liard Lowland. Between these two basins is
found a mildly elevated, but extensive and deeply incised mountain group known
as the Skeena Mountains.

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past glacier dynamics in the Cordillera. Because of
the irregular arrangement of valleys and ridges that forms the ice sheet’s
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them. In terms of
numerical glacier modelling, it implies complex patterns of flow and stress
within the ice, and requires high-resolution, and thus, large, computational
grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation over the model domain.
Due to the large latitudinal extent of the region, mean annual temperatures
vary between -12 and 12\,\degC. About half the model domain has a mean annual
temperature below freezing. However, seasonality, here computed as the
temperature difference between the warmest and the coldest months, varies
between around 10\,\degC along most of the Pacific Coast to above 40\,\degC in
the polar interior. This concerns most of the area above 60\degree~N, and
high-elevated mountain ranges below this latitude. There exist a partial
correlation between cold regions and those that have a large seasonality, so
that many regions where mean annual temperatures are well below freezing point
do experience warm summers.

Annual precipitation varies between less than half-a-meter in some of the
interior areas to several meters in some parts of maritime Alaska.
Precipitation rates are highest on the western slopes of the Pacific Coast
ranges and decline quickly on the other side of the mountains. In terms of ice
sheet mass balance, this contrast is even reinforced by the difference in
timing of the precipitation peak through the year. Coastal regions experience
most precipitation during winter and fall, thus as snow. On the contrary, many
inland regions experience dry winters and most of the annual precipitation
consequently falls as rain during the ablation season. The climatic setting of
the North American Cordillera is described further in \CCLI, where its effects
on the ice sheet model have been studied in detail.

@@

The Cordilleran ice sheet covered most of the Canadian Cordillera, as well
ajacent some of the adjacent land in north-western contiguous USA and Alaska.
The southern margin is the one that has been most studied. It consisted of
different ice lobes, that have formed at the exutory of the main valleys. The
Puget lobe was the westernmost and largest of these lobes. It covered the Puget
Lowland, a region that includes the current location of the cities of Vancouver
and Seattle. To the East of the Cascade Range were found the Okanagan and
Purcell lobes. The latter once dammed the Clark Fork River to form Glacial Lake
Missoula. As the ice sheet retreeted, the lake repeatedly and catastrophically
drained to the Pacific Ocean, forming on curse of the flood a lanscape now
known as the Channeled Scablands.

The western margin of the ice sheet covered part of the continental shelf. An
ice cap covering Vancouver Island was well-connected to main body of ice on the
continent. A large portion of the ice flow thus has engulfed between Vancouver
Island and the Olympic Peninsula, where it could find an exutory to the Pacific
Ocean, at the Juan the Fuca Lobe. Further to the North, the ice margin reached
Queen Charlotte Islands, yet the archipelago may not have been covered by ice
flowing from the land. Submarine troughs likely idicate the erosional action of
former ice streams, yet it is not clear wether they have formed during the last
glacial cycle or earlier.

The northern margin of the ice sheet is in several ways more complex than the
southern margin. Firstly geomorphological evidence indicate multiple glacier
advances. During the last glacial cycle, at least two glaciations occured, and
in some places, the first glaciation was more extensive than the second.
Secondly, geomorphological evidence indicate that the extent of glaciers in the
North was relatively restricted, despite very cold temperatures at present.

Finally, the eastern margin of the ice sheet is the most unclear. Although it
is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
the precise location of this junction is unknown, and little information exists
on its timing. More importantly, the nature of the interaction between the two
ice sheets is uncertain. It is not known for instance, if the junction lasted
long enough for the ice sheets to significantly alter their respective flow
dynamics. Because the present thesis focuses specifically on the Cordilleran
ice sheet, certainly our largest assumption consists in neglecting butressing
effects from the Laurentide ice sheet.

Although the last glacial maximum extent of the Cordilleran ice sheet, apart of
its eastern margin, is relatively well known, the history of deglaciation is
heavily debated, while little information exists on earlier dynamics of ice
sheet.
